Regularised parameter selection in the non-stationary Hawkes process

### 2017-04-27 _paper_output/paper_sparse_hawkes.tex
Explicate

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying a sparse driving
signal by means of \(\ell_1\) penalised regression where, by assumption,
the driving noise has a different form. This is analaogous to a blind
deconvolution problem, robust regression, or a compressive sensing-style
inverse problem, but considerably more challenging; While the process
model superfcially resembles a linear system, there are few of the
convenient results that would apply in a linear regression problem; and
as a time series problem it is difficult to choose our penalty
parameters by standard non-parametric means such as cross validation.
The proposed method uses a heuristically-motivated degrees-of-freedom
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

@@

We write the time series either by its values
\(\mathbf{t}=\{t_1, t_2,\dots,t_N(T)\}\). We represent the time series
either by its values or its arrival times as notationally conveneient.
We call it a (TODO) predictable Poisson point process if it has an
associated \emph{intensity} process \(\lambda\), such that \(\lambda\)
is predictable with respect to the filtration generated by \(N\), i.e.
\cite{daley_introduction_2003}

\[\lambda(\mathcal{F(t^-)}):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]

    The Hawkes process \cite{hawkes_point_1971} in particular, is the
linear, self-exciting point process; Concretely, given a "background"
rate \(\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,\) and a non-negative
influence kernel with positive support
\(\phi:\mathbb{R}\rightarrow\mathbb{R}^+\cap\{0\}\), the Hawkes process
has the following expression for its intensity:

@@

Here we follow a standard "thinning" method of simulating Hawkes
processes, given as Algorithm 7.5.IV of Daley and Vere-Jones'
\emph{Introduction to the Theory of Point Processes Volume 1}
\cite{daley_introduction_2003}, restating Ogata's original Algorithm 2
(\cite{ogata_lewis_1981}.

Since that is a general case, here we present a specific version for
linear Hawkes processes which is considerably simpler.

@@

Now the families
\(\mathrm{descendants}(t_j)\forall t_j\in \mathrm{children}(a_i)\) are
all once again identically and mutually independently distributed.
Proceeding by induction, it suffices to simulate the children of a
single point at time \(0\). The same method will suffice for its
children, children of children and so on. The final realisation can be
constructed from these simualtions by appropriate concatenation and
translation of the generations of each family.

The single kernel simulation is simple, since the Poisson intensity
process is no longer conditional on the history of the sequence. As
before, the influence kernel is \(\phi\). We will assume that \(\phi\)
is a.e. continuous \(L_1\) integrable and eventually decreasing in
\(t\). We associate with \(\phi\) a "majorant", \(\phi^{**}\) such that
\[\phi^{**}(t)\geq\phi(t), \forall t>0\] and \(\phi^{**}\) is
non-increasing and integrable. With the unimodal influence kernels used
in the current paper, such majorants are trivial to construct.

>>>>

We are interested in the problem of simultaneous identification of the
parameters of a Hawkes point process and identifying an inhomogeneous
driving signal by means of \(\ell_1\) penalised regression. This is
analaogous to a blind deconvolution problem \cite{stockham_blind_1975},
adaptive robust regression \cite{donoho_automatic_1988}, or a sparse
inverse problem \cite{tibshirani_regression_1996}, or, but considerably
more challenging; While the process model superficially resembles a
linear system, there are few of the convenient results that would apply
in a linear regression problem; and as a time series problem it is
difficult to choose our penalty parameters by standard non-parametric
means such as cross validation. The proposed method uses a
heuristically-motivated degrees-of-freedom regularisation parameter
selection, which we test through simulation. The proposed method works
well in practice.

@@

\subsubsection{Notation}\label{notation}

@@

We represent a realisation of the time series by its arrival times
\(\mathbf{t}=\{t_1, t_2,\dots,t_N(T)\}\). We may also, when as
notationally conveneient, use the equivalent representation as the
function that counts the arrivals:

\[
N(t):= \left|\{s:s\in \mathbf{t}\,\&\,s\leq t\}\right|
\]

We use only the left-continuous filtration
\(\mathcal{F}(\{N(s)\}_{s < t})\) generated by the process, which we
here write as \(\mathcal{F}(N)\)

We call it a (TODO) predictable Poisson point process if it has an
associated \emph{intensity} process \(\lambda\), such that \(\lambda\)
is predictable with respect to \(\mathcal{F}(N)\) generated by the
process, i.e.

\[\lambda(\mathcal{F}(t)):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]
\cite{daley_introduction_2003}

    The Hawkes process \cite{hawkes_point_1971} in particular, is the
linear, self-exciting point process; Concretely, given a "background"
rate \(\mu:\mathbb{R}^+\rightarrow\mathbb{R}^+,\) and a non-negative
influence kernel with positive support
\(\phi:\mathbb{R}\rightarrow\mathbb{R}^+\cup\{0\}\), the Hawkes process
has the following expression for its intensity:

@@

Here we follow a standard "thinning" method of simulating Hawkes
processes, specialised from Algorithm 7.5.IV of Daley and Vere-Jones'
\emph{Introduction to the Theory of Point Processes Volume 1}
\cite{daley_introduction_2003}, restating Ogata's original Algorithm 2
(\cite{ogata_lewis_1981}.

Since that is a general case, here we present, informally, a simpler,
specific version for linear Hawkes processes.

@@

Now the families
\(\mathrm{descendants}(t_j)\forall t_j\in \mathrm{children}(a_i)\) are
all once again identically and mutually independently distributed.
Proceeding by induction, it suffices to simulate the children of a
single point at time \(0\). The same method will suffice for its
children, children of children and so on. The final realisation can be
constructed from these simulations by appropriate concatenation and
translation of the generations of each family.

The single kernel simulation is simple, since the Poisson intensity
process is no longer conditional on the history of the sequence. As
before, the influence kernel is \(\phi\). We will assume that \(\phi\)
is a.e. continuous \(L_1\) integrable and eventually decreasing in
\(t\). We associate with \(\phi\) a "majorant", \(\phi^{**}\) such that
\[\phi^{**}(t)\geq\phi(t), \forall t>0\] and \(\phi^{**}\) is
non-increasing and integrable. As an aside, we mention that for the
unimodal influence kernels used in the current paper, such majorants are
trivial to construct.
